Forecasting Gamma-Ray Bursts using Gravitational Waves

### 2018-04-22 Paper/Manuscript.tex
Niels's corrections, a few more references.

\begin{abstract}
We explore the intriguing possibility of employing future ground-based gravitational-wave interferometers to detect the inspiral of binary neutron stars sufficiently
early to alert electromagnetic observatories so that a gamma-ray burst (GRB) can be observed in its entirety from its very beginning.
We quantify the ability to predict a GRB by computing the time a binary neutron star (BNS) system takes to inspiral from its moment of detection to its final merger. We define the moment of detection to be the instant at which the inteferometer network accumulates a signal-to-noise ratio of 15. %for the BNS inspiral.
For our computations, %of advance warning times, 
we specifically consider BNS systems at luminosity distances of (i) $D\le200\,$Mpc in the three-interferometer Advanced-LIGO-Virgo network of 2020, and (ii) $D \le 1000\,$Mpc in the Einstein Telescope's B and C configurations. 
In the case of Advanced LIGO-Virgo we find that we may at best get a few minutes of warning time, thus we expect no forecast of GRBs in the 2020s. 
On the other hand, Einstein Telescope will provide us with advance warning times of up to five hours.  
Taking one hour as a benchmark warning time, we obtain a corresponding horizon
distance of roughly 600 Mpc for the Einstein Telescope C configuration.
%, more specifically, greater than an hour
%for BNS inspirals closer than roughly 600\,Mpc. Taking this luminosity distance to be our horizon distance, 
Using current BNS event rates within this horizon distance, we show that this translates to forecasts of $\gtrsim \ord(10^2)$ GRBs in the 2030s. %with Einstein C. %with the C configuration of Einstein Telescope.
We reapply our warning-time computation to binary black hole - neutron star inspirals and find that we expect 1 to 3 tidal disruption events 
to be forecast by the same detector.
%Our main conclusions are: (i) we expect ALV network to forecast 1 to 3 GRBs in the 2020s, (ii)

\end{abstract}
\maketitle

@@

GW170817 was initially identified by the LIGO-Hanford interferometer (H1) 
using a template bank of gravitational waveforms computed within the framework of post-Newtonian theory \cite{Buonanno:2009zt, Blanchet_LRR}.
The inspiral swept across the Advanced Ligo-Virgo (ALV) network's bandwidth from approximately $ 30\,$Hz to about $ 2\,$kHz in 57 seconds,
executing nearly $ 3000$ gravitational wave cycles and accumulating a network signal-to-noise ratio of 32.4 with a false-alarm rate of one per $8.0\times 10^4$ years \cite{GW170817}. 
The total mass of the system was inferred to be $2.74^{+0.04}_{-0.01} M_\odot$ with component masses in the range $1.17-1.60 M_\odot$, consistent
with neutron stars (assuming low spin priors, see Sec.~\ref{sec:idealizations}).
Moreover, both the gravitational wave (GW) and EM observations were consistent with a source luminosity distance of $\sim\!\! 40\,$Mpc
and supported the hypothesis that GW170817 resulted from the inspiral and merger of two neutron stars, 
causing the prompt short-hard gamma-ray burst (GRB170817A) in NGC4993 as first suggested by Ref.~\cite{Eichler:1989ve}.

@@

The short answer is in the affirmative. The long answer is still affirmative, but depends on many factors such as source distance and sky position, detector sensitivity and orientation. One also needs to account for the fact that GRBs are thought to be collimated emissions with narrow outburst angles \cite{Kumar:2014upa}.
As such, most GRBs that occur in the universe are electromagnetically undetectable. Fortunately, this was not the case with GRB170817A, 
but even for a reasonable GRB jet opening angle of $\sim 10^\circ$, we expect to observe $\sim 10\%$ of neutron star mergers as GRBs \cite{Patricelli:2016bkt}. 
However, it is possible that some of the near-Earth pointing short hard GRBs may be observed as long-duration GRBs or X-ray flashes thus increasing this percentage \cite{Bucciantini:2011kx}.

@@

Our computation for $T_\text{AW}$ is based on the Newtonian evolution of a binary system composed of two point masses in a quasi-circular Keplerian orbit. 
The long-term dynamics is governed by the radiation reaction of 
the GW emission from the system due to the time variance of the quadrupole moment of the binary.
Though the Newtonian (leading-order) treatment gives the 
inspiral time with better than $ 95\%$ accuracy,
we nonetheless supplement it with post-Newtonian corrections up to 3.5 post-Newtonian order (3.5PN) \cite{Blanchet_LRR}.
However, even that fails to provide an accurate depiction of the binary
evolution in the strong-field regime, where speeds $\gtrsim 0.1c$. This is a very complex and complicated problem requiring fully
numerical treatment formulated in terms of general relativistic, magnetohydrodynamic codes 
with neutrino transport running on large-scale computing clusters and taking up to millions of CPU hours 
(see Ref.~\cite{Faber:2012rw} for a review, and Refs.~\cite{Kyutoku:2017voj, Zappa:2017xba, Dietrich:2018upm} for the latest developments and references therein).
Despite its failures, our quadrupole radiation driven Newtonian
evolution is more than sufficient to provide reliable estimations of advance warning times.
We support this claim in Sec.~\ref{sec:idealizations} with an extensive list of computations showing how much each of our approximations affects the inspiral time.

Let us emphasize that this article is intended as a pedagogical introduction to gravitational-wave astronomy written at a level accessible to
Ph.D. students, advanced undergraduates, and colleagues in astronomy and/or astrophysics who wish to learn more about the underlying physics.
Be that as it may, even the experts might find our results motivating and exciting.

This article is organized as follows. Sec.~\ref{sec:BNS_inspiral} introduces the formulation for the leading-order evolution of the binary inspiral.
Sec.~\ref{sec:IFO_response} details the interferometer response to GWs based on topology.
Sec.~\ref{sec:idealizations} lists the various idealizations we employ to simplify our treatment and supplies justification for each.
Sec.~\ref{sec:GW170817} tests our model network and evolution using the parameters of GW170817 and the corresponding observations.
Sec.~\ref{sec:results} contains our main results presented as advance warning times in Tables~\ref{table:LIGO2020} and \ref{table:ET}; 
and as horizon distances and event rates in Table~\ref{table:horizon}.
In Sec.~\ref{sec:BH_NS}, we recompute the advance warning times in the case of black hole - neutron star inspirals to see whether or not 
future detectors can forecast a tidal disruption event, for which we summarize our findings in Table~\ref{table:ET_BH_NS}.

Throughout the text, $f$ denotes the GW frequency of the dominant quadrupole mode, which is twice the orbital (Keplerian) frequency $f_K$. 
%Relatedly, we define the following angular frequencies: $\omega = 2\pi f, \Omega=2\pi f_K$ with $\omega=2\Omega$.
We employ the $\simeq$ symbol when displaying the numerical values of quantities %with infinite decimal expansions 
which we usually truncate at four significant digits.
The $\approx$ symbol is reserved for approximations whereas $\sim$ denotes rough, order-of-magnitude equalities,
e.g., $\pi \simeq 3.142$, $\pi \approx 3$, $\pi \sim \ord(1)$. %Kepler's third law: $f_K^2\sim r^{-3}$.
Overdots denote time derivatives with respect to detector-frame time, e.g., $\dot{E} =dE/dt$, and $\propto$ denotes proportionalities. 
Unless otherwise noted, we use standard SI units.

\section{Binary neutron star inspirals}\label{sec:BNS_inspiral}
\subsection{Newtonian evolution}
For us, the starting point for the evolution of a binary under gravitational radiation reaction
%the emission of GWs 
is the expression for the power emitted in GWs given by 
the celebrated Einstein quadrupole formula \cite{1918SPAW}
%for the generated by the time-variance of the mass quadrupole moment
%
\be
\dot{E} = \f{G}{5c^5}\langle \dddot{Q}_{ij}\dddot{Q}_{ij} \rangle \label{eq:Edot_quadrupole}\, .
\ee
%
Here, $Q_{ij}$ is a trace-reversed {\bf MAYBE: explain this} mass quadrupole moment which
is the dominant contribution to the power emitted %in gravitational radiation
by a system with changing mass moments. Details of how to obtain this expression can be found in Ref.~\cite{Maggiore}.

@@

An important quantity in GW astronomy is the chirp mass of the binary given by
%
\be
M_c = \mu^{3/5} M^{2/5} = \f{(m_1 m_2)^{3/5}}{(m_1+m_2)^{1/5}} \label{eq:chirp_mass}\, ,
\ee
%
which turns Eq.~(\ref{eq:Edot}) into
%
\be
\dot{E} = \f{32}{5}\f{c^5}{G} \left(\f{G M_c\, \omega}{2c^3}\right)^{10/3}, \label{eq:Edot2}
\ee
%
where $\omega=2\Omega$ is the angular frequency of the quadrupolar GWs. 
The quantity in parantheses in Eq.~(\ref{eq:Edot2}) equals $x^{3/2}$ where 
$x \equiv ({G M_c\, \Omega}/{c^3})^{2/3} $
is the dimensionless inverse separation parameter ubiquitous in post-Newtonian theory.
As $x\propto v^2/c^2$, it is used to keep track of post-Newtonian orders and we can see from Eq.~(\ref{eq:Edot2}) that $\dot{E}\propto x^5$. 
%hence the common statement in the literature saying that gravitational radiation starts

The emission of power in GWs decreases the total energy of the binary given by
%
\be
E_b = -\frac{G m_1 m_2}{2r}\,.  \label{eq:binding_energy1}
\ee
%
$\dot{E}_b \propto \dot{r}/r^2$, so $\dot{E}_b < 0$ implies $\dot{r}<0$, i.e, the orbital separation decreases
making the orbit more bound in time.
Using Kepler's third law, $ \Omega^2 = G M r^{-3}$, and Eq.~(\ref{eq:chirp_mass}) we obtain
%
\be
E_b = -\left(\f{G^2 M_c^5 \omega^2}{32}\right)^{1/3}\, . \label{eq:binding_energy2}
\ee
%
Energy conservation dictates that $\dot{E}=-\dot{E}_b$. Using the chain rule to determine $\dot{E}_b$ from Eq.~(\ref{eq:binding_energy2})
and setting the resulting expression equal to Eq.~(\ref{eq:Edot2}) yields $\dot{\omega}$ as a function of $\omega$ and various constants.
Translating this expression using $\omega=2\pi f$ we arrive at the standard expression for the GW frequency evolution
%%
%
\be
\dot{f} = \f{96}{5}\pi^{8/3} \f{(G M_c)}{c^5}^{5/3}\, f^{11/3} \label{eq:fdot}
\ee
%
This can be integrated straightforwardly after defining a new time variable $\tau\equiv t_\text{coal}-t$ that equals zero when the neutron stars coalesce.
Using $\tau$ we can now define an inspiral time at a given GW frequency
%
\be
\tau_\text{insp}(f) = \f{5}{256\pi}\f{c^5}{(\pi G M_c)^{5/3}} \,f^{-8/3}\label{eq:tau_insp}\, ,
\ee
%
%
%\be
%\tau_\text{insp}(f) \equiv \f{5}{256\pi}\f{c^5}{(\pi G M_c)^{5/3}}\left[ f^{-8/3}-f_\text{ISCO}^{-8/3}\right]\label{eq:tau_insp} \, .
%\ee
%
which we can rewrite as follows
%
\be
\tau_\text{insp}(f) \simeq 16.72\,\text{minutes} \, \left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\,\left(\f{10\,\text{Hz}}{f}\right)^{8/3}, \label{eq:tau_insp2}
\ee
%
where $1.219 M_\odot$ is the chirp mass corresponding to $m_1=m_2=1.4 M_\odot$.

We can also compute the number of GW cycles over the course of an inspiral. Given that the orbital period $T_K(t)$ varies over a time scale much larger than $T_K$ itself
we can write
%
\be
\mathcal{N}_\text{cyc}(f) = \int_f^{f_\text{max}} \f{f'}{\dot{f'}}\, df' =\f{1}{32\pi} \f{c^5}{(\pi G M_c)^{5/3}}\left[ f^{-5/3}-f_\text{max}^{-5/3}\right] \label{eq:Ncyc1}
\ee
%
We now introduce a cut-off for the inspiral imposed by the frequency of the innermost stable circular (ISCO) orbit in Schwarzschild spacetime
\be
\Omega_\text{ISCO} = \f{c^3}{6^{3/2} G M} \label{eq:Sch_f_isco}.
\ee
%
%
We will see in Sec.~\ref{sec:idealizations} that $f_\text{ISCO} \sim 1000\,$Hz for a system with $M\approx 2 M_\odot$. 
On the other hand, the frequencies of interest for our advance warning time estimations will be $\lesssim 10\,$Hz
meaning that $f^{-5/3} \gg f_\text{ISCO}^{-5/3}$. Therefore, at the leading (Newtonian) order, the number of GW cycles can be approximated by
%
\be
\mathcal{N}_\text{cyc}(f) \approx \f{1}{32\pi} \f{c^5}{(\pi G M_c)^{5/3}}\, f^{-5/3} \,  \label{eq:Ncyc2}
\ee
%
%Restoring some familiar numbers in this expression yields
%
which can be rewritten as
\be
\mathcal{N}_\text{cyc}(f)\approx 1.605\times 10^4 \left(\f{10\,\text{Hz}}{f} \right)^{5/3}\,\left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\label{eq:Ncyc3}\, .
\ee
%
This expression is quite telling: if a ground-based interferometer picks up an
inspiralling BNS at $f=10\,$Hz then there will be $\gtrsim \ord(10^4)$ GW cycles in the detector's data stream until the merger.

@@

Let us now turn our attention to the GWs generated by the inspiral.
These are tensorial perturbations propagating at the speed of light in flat (Minkowski) spacetime.
In the so-called wave zone, defined by the condition $ \lambda= c/f \ll D$, where $\lambda$ is the GW wavelength and $D$ is the distance to the GW source, 
these tensor modes satisfy the flat spacetime sourceless wave equation. %$\Box\bar{h}_{\mu\nu} = 0$.
%meaning that we are considering the solutions far from the GW source, 
In a suitable gauge, such as the commonly used transverse-traceless gauge, 
it can be shown that there exist only two physical degrees of freedom which manifest themselves as two independent polarization amplitudes $h_+$ and $h_\times$. %when interacting with masses away from the source.
For a binary in a circular orbit, these states read
%
\begin{align}
 h_+(t) &= h_c(t)\, \left(\f{1+\cos^2\iota}{2}\right)\, \cos[\Phi_\text{N}(t)],\label{eq:hplus_TD}\\
 h_\times(t) & =h_c(t)\,\cos\iota \sin[\Phi_\text{N}(t)]\label{eq:hcross_TD},
\end{align}
%
%
where
%
\be
h_c(t) =\f{4}{D}\left(\f{G M_c}{c^2}\right)^{5/3}\, \left(\f{\pi f(t)}{c}\right)^{2/3} \label{eq:h_c}
\ee
is the characteristic strain and
$\iota=\cos^{-1}(\mathbf{\hat{n}}\cdot{\mathbf{\hat{L}}}) $ is the inclination angle between the line of sight unit vector $\mathbf{\hat{n}} $ and the orbital angular momentum unit vector $\mathbf{\hat{L}}$. 
For an orbit seen edge on, $\iota = \pi/2$ giving $h_\times=0$, i.e,  a purely plus-polarized wave.
When $\iota=0$ the orbit is seen face-on and we have a circularly-polarized wave: $\langle h_+ \rangle = \langle h_\times\rangle$ where $\langle \ldots \rangle$
denote time-averages over one orbit. Note that as $f$ increases in time, so does $h(t)$, hence the characteristic ``chirping'' of GW signals.

@@

for $M \equiv m_1+m_2 = 2.8 M_\odot$. The true ISCO is located at $r< r_\text{ISCO}$ with frequency possibly up to 1.7\,kHz \cite{Marronetti:2003hx}, 
however, there is no simple expression for it akin to Eq.~(\ref{eq:f_isco}) and its extraction from numerical simulations is a rather involved procedure.

>>>>

\begin{abstract}
We explore the intriguing possibility of employing future ground-based gravitational-wave interferometers to detect the inspiral of binary neutron stars sufficiently
early to alert electromagnetic observatories so that a gamma-ray burst (GRB) can be observed in its entirety from its very beginning.
We quantify the ability to predict a GRB by computing the time a binary neutron star (BNS) system takes to inspiral from its moment of detection to its final merger. We define the moment of detection to be the instant at which the inteferometer network accumulates a signal-to-noise ratio of 15. %for the BNS inspiral.
For our computations, %of advance warning times, 
we specifically consider BNS systems at luminosity distances of (i) $D\le200\,$Mpc in the three-interferometer Advanced-LIGO-Virgo network of 2020, and (ii) $D \le 1000\,$Mpc in the Einstein Telescope's B and C configurations. 
In the case of Advanced LIGO-Virgo we find that we may at best get a few minutes of warning time, thus we expect no forecast of GRBs in the 2020s. 
On the other hand, Einstein Telescope will provide us with advance warning times of up to five hours.  
Taking one hour as a benchmark warning time, we obtain a corresponding horizon
distance of roughly 600 Mpc for the Einstein Telescope C configuration.
%, more specifically, greater than an hour
%for BNS inspirals closer than roughly 600\,Mpc. Taking this luminosity distance to be our horizon distance, 
Using current BNS event rates within this horizon distance, we show that this translates to forecasts of $\gtrsim \ord(10^2)$ GRBs in the 2030s. %with Einstein C. %with the C configuration of Einstein Telescope.
We reapply our warning-time computation to binary black hole - neutron star inspirals and find that we expect 1 to 3 tidal disruption events 
to be forecast by the same detector.
This article is intended as a pedagogical introduction to gravitational-wave astronomy written at a level accessible to
Ph.D. students, advanced undergraduates, and colleagues in astronomy and/or astrophysics who wish to learn more about the underlying physics.
Though many of our results may be known to the experts, they might nonetheless find this article motivating and exciting.
\end{abstract}
\maketitle

@@

GW170817 was initially identified by the LIGO-Hanford interferometer (H1) 
using a template bank of gravitational waveforms computed within the framework of post-Newtonian theory \cite{Buonanno:2009zt, Blanchet_LRR}.
The inspiral swept across the Advanced Ligo-Virgo (ALV) network's bandwidth from approximately $ 30\,$Hz to about $ 2\,$kHz in 57 seconds,
executing nearly $ 3000$ gravitational wave cycles and accumulating a network signal-to-noise ratio of 32.4 with a false-alarm rate of one per $8.0\times 10^4$ years \cite{GW170817}. 
The total mass of the system was inferred to be $2.74^{+0.04}_{-0.01} M_\odot$ with component masses in the range $1.17-1.60 M_\odot$, consistent
with neutron stars (assuming low spin priors, see Sec.~\ref{sec:idealizations}).
Moreover, both the gravitational wave (GW) and EM observations were consistent with a source location in the galaxy in NGC4993 
at a luminosity distance of $\sim\!\! 40\,$Mpc,
and supported the hypothesis that GW170817 resulted from the inspiral and merger of two neutron stars, 
causing a prompt short-hard gamma-ray burst (GRB170817A) as first suggested by Ref.~\cite{Eichler:1989ve}.

@@

The short answer is ``yes''. The long answer is still ``yes'', but depends on many factors such as source distance and sky position, detector sensitivity and orientation. One also needs to account for the fact that GRBs are thought to be collimated emissions with narrow outburst angles \cite{Kumar:2014upa}.
As such, most GRBs that occur in the universe are electromagnetically undetectable. Fortunately, this was not the case with GRB170817A, 
but even for a reasonable GRB jet opening angle of $\sim 10^\circ$, we expect to observe $\sim 10\%$ of neutron star mergers as GRBs \cite{Patricelli:2016bkt}. 
However, it is possible that some of the near-Earth pointing short hard GRBs may be observed as long-duration GRBs or X-ray flashes thus increasing this percentage \cite{Bucciantini:2011kx}.

@@

Our computation for $T_\text{AW}$ is based on the Newtonian evolution of a binary system composed of two point masses in a quasi-circular Keplerian orbit. 
The long-term dynamics is governed by the radiation reaction of 
the GW emission from the system due to the time variance of the quadrupole moment of the binary.
Though the Newtonian (leading-order) treatment gives the 
inspiral time with better than $ 95\%$ accuracy,
we nonetheless supplement it with post-Newtonian corrections up to 3.5 post-Newtonian order (3.5PN) \cite{Blanchet_LRR}.
However, even that fails to provide an accurate depiction of the binary
evolution in the strong-field regime, where speeds $\gtrsim 0.1c$. This is a very complex and complicated problem requiring fully
numerical treatment formulated in terms of general relativistic, magnetohydrodynamic codes 
with neutrino transport running on large-scale computing clusters and taking up to millions of CPU hours 
(see Ref.~\cite{Faber:2012rw} for a review, and Refs.~\cite{Baiotti:2016qnr,Kyutoku:2017voj, Zappa:2017xba, Dietrich:2018upm} for the latest developments and references therein).
Despite its simple formulation, our quadrupole radiation driven Newtonian
evolution is more than sufficient to provide reliable estimations of advance warning times.
We support this claim in Sec.~\ref{sec:idealizations} with an extensive list of computations showing how much each of our approximations affects the inspiral time.

%Let us emphasize that this article is intended as a pedagogical introduction to gravitational-wave astronomy written at a level accessible to
%Ph.D. students, advanced undergraduates, and colleagues in astronomy and/or astrophysics who wish to learn more about the underlying physics.
%Be that as it may, even the experts might find our results motivating and exciting.
%
This article is organized as follows. Sec.~\ref{sec:BNS_inspiral} introduces the formulation for the leading-order evolution of the binary inspiral.
Sec.~\ref{sec:IFO_response} details the interferometer response to GWs based on topology.
Sec.~\ref{sec:idealizations} lists the various idealizations we employ to simplify our treatment and supplies justification for each.
Sec.~\ref{sec:GW170817} tests our model network and evolution using the parameters of GW170817 and the corresponding observations.
Sec.~\ref{sec:results} contains our main results presented as advance warning times in Tables~\ref{table:LIGO2020} and \ref{table:ET}; 
and as horizon distances and event rates in Table~\ref{table:horizon}.
In Sec.~\ref{sec:BH_NS}, we recompute the advance warning times in the case of black hole - neutron star inspirals to see whether or not 
future detectors can forecast a tidal disruption event, for which we summarize our findings in Table~\ref{table:ET_BH_NS}.

Throughout the text, $t$ denotes observer/detector time and $f$ denotes the GW frequency of the dominant quadrupole mode, which is twice the orbital (Keplerian) frequency $f_K$. 
%Relatedly, we define the following angular frequencies: $\omega = 2\pi f, \Omega=2\pi f_K$ with $\omega=2\Omega$.
We employ the $\simeq$ symbol when displaying the numerical values of quantities %with infinite decimal expansions 
which we usually truncate at four significant digits.
The $\approx$ symbol is reserved for approximations whereas $\sim$ denotes rough, order-of-magnitude equalities,
e.g., $\pi \simeq 3.142$, $\pi \approx 3$, $\pi \sim \ord(1)$. %Kepler's third law: $f_K^2\sim r^{-3}$.
Overdots denote time derivatives with respect to detector-frame time, e.g., $\dot{E} =dE/dt$, and $\propto$ denotes proportionalities. 
Unless otherwise noted, we use standard SI units.

\section{Binary neutron star inspirals}\label{sec:BNS_inspiral}
\subsection{Newtonian evolution}
For us, the starting point for the evolution of a binary under gravitational radiation reaction
%the emission of GWs 
is the expression for the power emitted in GWs given by 
the celebrated Einstein quadrupole formula \cite{1918SPAW}
%for the generated by the time-variance of the mass quadrupole moment
%
\be
\dot{E} = \f{G}{5c^5}\langle \dddot{Q}_{ij}\dddot{Q}_{ij} \rangle \label{eq:Edot_quadrupole}\, .
\ee
%
Here, $Q_{ij}$ is a trace-reversed mass quadrupole moment which
is the dominant contribution to the power emitted %in gravitational radiation
by a system with changing mass moments. Details of how to obtain this expression can be found in Ref.~\cite{Maggiore}.

@@

An important quantity in GW astronomy is the chirp mass of the binary given by
%
\be
M_c = \mu^{3/5} M^{2/5} = \f{(m_1 m_2)^{3/5}}{(m_1+m_2)^{1/5}} \label{eq:chirp_mass}\, ,
\ee
%
which turns Eq.~(\ref{eq:Edot}) into
%
\be
\dot{E} = \f{32}{5}\f{c^5}{G} \left(\f{G M_c\, \omega}{2c^3}\right)^{10/3}= \f{32}{5}\f{c^5}{G}x^5, \label{eq:Edot2}
\ee
%
where $x = ({G M_c\, \Omega}/{c^3})^{2/3} $ is the dimensionless inverse separation parameter ubiquitous in post- Newtonian theory and $\omega=2\Omega$ is the angular frequency of the quadrupolar GWs. 
As $x\propto v^2/c^2$, it is used to keep track of post-Newtonian orders and we can see from Eq.~(\ref{eq:Edot2}) that $\dot{E}\propto x^5$. 
%hence the common statement in the literature saying that gravitational radiation starts

The emission of power in GWs decreases the total energy of the binary given by
%
\be
E_b = -\frac{G m_1 m_2}{2r}\,.  \label{eq:binding_energy1}
\ee
%
$\dot{E}_b \propto \dot{r}/r^2$, so $\dot{E}_b < 0$ implies $\dot{r}<0$, i.e, the orbital separation decreases
making the orbit more bound in time.
Using Kepler's third law, $ \Omega^2 = G M r^{-3}$, and Eq.~(\ref{eq:chirp_mass}) we obtain
%
\be
E_b = -\left(\f{G^2 M_c^5 \omega^2}{32}\right)^{1/3}\, . \label{eq:binding_energy2}
\ee
%
Energy conservation dictates that $\dot{E}=-\dot{E}_b$. Using the chain rule to determine $\dot{E}_b$ from Eq.~(\ref{eq:binding_energy2})
and setting the resulting expression equal to Eq.~(\ref{eq:Edot2}) yields $\dot{\omega}$ as a function of $\omega$ and various constants.
Translating this expression using $\omega=2\pi f$ we arrive at the standard expression for the GW frequency evolution
%%
%
\be
\dot{f} = \f{96}{5}\pi^{8/3} \f{(G M_c)}{c^5}^{5/3}\, f^{11/3} \label{eq:fdot}
\ee
%
This can be integrated straightforwardly after defining a new time variable $\tau\equiv t_\text{coal}-t$ that equals zero when the neutron stars coalesce.
Using $\tau$ we can now define an inspiral time at a given GW frequency
%
\be
\tau_\text{insp}(f) = \f{5}{256\pi}\f{c^5}{(\pi G M_c)^{5/3}} \,f^{-8/3}\label{eq:tau_insp}\, ,
\ee
%
%
%\be
%\tau_\text{insp}(f) \equiv \f{5}{256\pi}\f{c^5}{(\pi G M_c)^{5/3}}\left[ f^{-8/3}-f_\text{ISCO}^{-8/3}\right]\label{eq:tau_insp} \, .
%\ee
%
which we can rewrite as follows
%
\be
\tau_\text{insp}(f) \simeq 16.72\,\text{minutes} \, \left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\,\left(\f{10\,\text{Hz}}{f}\right)^{8/3}, \label{eq:tau_insp2}
\ee
%
where $1.219 M_\odot$ is the chirp mass corresponding to $m_1=m_2=1.4 M_\odot$. This is the canonical value for the 
mass of a neutron star due to Chandrasekhar limit \cite{Chandra}.

We can also compute the number of GW cycles over the course of an inspiral. Given that the orbital period $T_K(t)$ varies over a time scale much larger than $T_K$ itself
we can write
%
\be
\mathcal{N}_\text{cyc}(f) = \int_f^{f_\text{max}} \f{f'}{\dot{f'}}\, df' =\f{1}{32\pi} \f{c^5}{(\pi G M_c)^{5/3}}\left[ f^{-5/3}-f_\text{max}^{-5/3}\right] \label{eq:Ncyc1}
\ee
%
We now introduce a cut-off for the inspiral imposed by the frequency of the innermost stable circular (ISCO) orbit in Schwarzschild spacetime
\be
\Omega_\text{ISCO} = \f{c^3}{6^{3/2} G M} \label{eq:Sch_f_isco}.
\ee
%
%
We will see in Sec.~\ref{sec:idealizations} that $f_\text{ISCO} \sim \ord(10^3)\,$Hz for a system with $M =2.8 M_\odot$. 
On the other hand, the frequencies of interest for our advance warning time estimations are $\lesssim 10\,$Hz
meaning that $f^{-5/3} \gg f_\text{ISCO}^{-5/3}$. Therefore, at the leading (Newtonian) order, the number of GW cycles can be approximated by
%
\be
\mathcal{N}_\text{cyc}(f) \approx \f{1}{32\pi} \f{c^5}{(\pi G M_c)^{5/3}}\, f^{-5/3} \,  \label{eq:Ncyc2}
\ee
%
%Restoring some familiar numbers in this expression yields
%
which can be rewritten as
\be
\mathcal{N}_\text{cyc}(f)\approx 1.605\times 10^4 \left(\f{10\,\text{Hz}}{f} \right)^{5/3}\,\left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\label{eq:Ncyc3}\, .
\ee
%
This expression is quite telling: if a ground-based interferometer picks up an
inspiralling BNS at $f=10\,$Hz then there will be $\gtrsim \ord(10^4)$ GW cycles in the detector's data stream until the merger.

@@

Let us now turn our attention to the GWs generated by the inspiral.
These are tensorial perturbations propagating at the speed of light in flat (Minkowski) spacetime.
In the so-called wave zone, defined by the condition $ \lambda= c/f \ll D$, where $\lambda$ is the GW wavelength and $D$ is the distance to the GW source, 
these tensor modes satisfy the flat spacetime sourceless wave equation. %$\Box\bar{h}_{\mu\nu} = 0$.
%meaning that we are considering the solutions far from the GW source, 
In a suitable gauge, such as the commonly used transverse-traceless gauge, 
it can be shown that there exist only two physical degrees of freedom which manifest themselves as two independent polarization amplitudes $h_+$ and $h_\times$. %when interacting with masses away from the source.
For a binary in a circular orbit, these states read \cite{Maggiore}
%
\begin{align}
 h_+(t) &= h_c(t)\, \left(\f{1+\cos^2\iota}{2}\right)\, \cos[\Phi_\text{N}(t)],\label{eq:hplus_TD}\\
 h_\times(t) & =h_c(t)\,\cos\iota \sin[\Phi_\text{N}(t)]\label{eq:hcross_TD},
\end{align}
%
%
where
%
\be
h_c(t) =\f{4}{D}\left(\f{G M_c}{c^2}\right)^{5/3}\, \left(\f{\pi f(t)}{c}\right)^{2/3} \label{eq:h_c}
\ee
is the characteristic strain and
$\iota=\cos^{-1}(\mathbf{\hat{n}}\cdot{\mathbf{\hat{L}}}) $ is the inclination angle between the line of sight unit vector $\mathbf{\hat{n}} $ and the orbital angular momentum unit vector $\mathbf{\hat{L}}$. 
For an orbit seen edge on, $\iota = \pi/2$ giving $h_\times=0$, i.e,  a purely plus-polarized wave.
When $\iota=0$ the orbit is seen face-on and we have a circularly-polarized wave: $\langle h_+ \rangle = \langle h_\times\rangle$ where $\langle \ldots \rangle$
denote time-averages over one orbit. Note that as $f$ increases in time, so does $h(t)$, hence the characteristic ``chirping'' of GW signals.

@@

for $M \equiv m_1+m_2 = 2.8 M_\odot$. In the strong-field, highly-dynamical spacetime of the last few orbits, the
concept of an ISCO becomes fuzzy due to strong dissipative effects. However, an ISCO can nonetheless be dynamically determined
with frequency possibly up to 1.7\,kHz \cite{Marronetti:2003hx}, 
but there is no simple expression for it akin to Eq.~(\ref{eq:f_isco}) and its extraction from numerical simulations is a rather involved procedure.
